Parallelizing Linear Recurrent Neural Nets Over Sequence Length

### 2017-10-27 paper/main.tex
More paper edits

\subsection{Throughput Benchmarks}
We first illustrate the throughput advantage of the parallel scan algorithm for
evaluating the linear recurrence. For a minibatch comprised of \(b\) sequences
of length \(T\), we define the number of events as \(bT\) and the throughput as
the number of events per second. We implement two CUDA kernels, one which
evaluates the parallel linear recurrence described in algorithm \ref{alg:plr},
and one which evaluates the linear recurrence using the standard serial
approach (with parallelization over features but not sequence length).
This comparison is performed directly at the kernel level, avoiding
any overhead from TensorFlow. We find that the parallel kernel has
a distinct advantage at long sequence lengths with a speedup factor of up to 40x,
as shown in table \ref{table:kernel-throughput}.
At short sequence lengths, the parallel kernel does not perform well due to the
overhead of multiple passes over data and communication between processors.
\begin{table}[]
  \centering
  \caption{
    Speedup of the parallel kernel compared to the serial kernel at various
    values of $(\text{minibatch size } \mathbf{x} \text{ features}) = m$.
    Performance was directly measured without any overhead from TensorFlow.
  }
\label{table:kernel-throughput}
\begin{tabular}{@{}llll@{}}
Sequence Length & \(m=4\)  & \(m=32\) & \(m=128\) \\ \midrule
16              & 0.06 & 0.06 & 0.05  \\
256             & 0.22 & 0.22 & 0.86  \\
4,096           & 1.02 & 2.94 & 3.36  \\
65,536          & 38.5 & 41.8 & 17.5  \\ \bottomrule
\end{tabular}
\end{table}

Many recently introduced RNNs can be accelerated with the parallel linear recurrence
algorithm.
Table \ref{table:rnn-throughput} shows the
throughput advantage by using a parallel linear recurrence when compared to a
serial linear recurrence. Since PLR and SLR are simply different methods for
computing the recurrence, there is no penalty incurred for this speedup - we
obtain exactly the same output.

Typically experiments will aim to maximize the memory
utilisation of the GPU, so we simulate this by choosing the product of the batch
size and the sequence length to be constant. We see that the architectures which
have fewer terms (for which the linear recurrence is a higher proportion of the
computational burden) are more effected by the switch to PLR.  This is
particularly clear in the case of the QRNN, where including a wider
convolutional window results in less time spent on the recurrence and less to
gain from the parallelization.
\begin{figure}
\begin{center}
  \begin{tabular}{@{}lrrrr@{}}
    \label{table:rnn-throughput}
Sequence Length & GILR-LSTM & SRU & QRNN(2) & QRNN(10)\\ \midrule
16 & 0.61 & 0.28 & 0.38 & 0.78\\
256 & 0.91 & 0.84 & 0.86 & 0.99\\
4,096 & 0.98 & 1.38 & 1.18 & 1.05\\
65,536 & 1.41 & 9.21 & 6.68 & 2.05\\ \bottomrule
  \end{tabular}
\end{center}
\caption{Speedup obtained when using parallel linear recurrence compared to
  serial linear recurrence. We analyse this for several recent architectures:
  the Simple Recurrent Unit from \cite{lei2017}, the Quasi-RNN from
  \cite{bradbury2017quasi}, with a convolutional filter of size 2 and 10, and
  the GILR-LSTM which we introduce in section \ref{sec:ls-rnns}. To give a fair
  model of a typical use-case maximising the memory usage of the GPU, we set
  batch size \(b\) such that \(Tb = 65536\) for a sequence length \(T\).}
\end{figure}

Finally, we compare the throughput of the linear surrogate LSTM to the current
fastest available implementation of the LSTM, the CuDNN-LSTM. Figure
\ref{fig:tp_perf} shows that our GILR-LSTM has higher throughput for most of the
low-batch-size regime, with a 30x speedup at a sequence length of 65,536. Recent
work \cite{keskar2017large} has suggested that small mini-batches may aid
generalization by encouraging the discovery of shallow minima. The GILR-LSTM
allows us to achieve high throughput for all choices of minibatch size and
sequence length.

@@

\begin{table}[t]
\centering
\includegraphics[width=12cm]{cudnn_heatmap.png}
\caption{Throughput comparison between LSTM-256-256 with GILR-LSTM-256-256, with a
  32-dimensional input and a 2-dimensional output, for various batch sizes and
  sequence lengths.The LSTM only has a single row of data because its throughput
  is independent of sequence length. Entries are missing from the GILR-LSTM table
  because there was not enough memory on the GPU to handle such large batch
  sizes and sequences.}
\label{fig:tp_perf}
\end{table}

@@

\subsection{Synthetic Example}
In order to demonstrate how PLR may be used to speed up tasks that the LSTM is
well-suited for, we tackle a pathological inference problem. Originally
introduced as example 2b from Hochreiter and Schmidhtbuer's introduction of the
LSTM, \cite{hochreiter1997long}, it involves storing a number for a very long
sequence of time steps. We have an alphabet of size \(p\) with each character
represented as a particular one-hot vector in this space. We input a sequence of
\(n\) of these vectors, chosen at random (with replacement). The first vector in
the sequence is always the same, up to the sign of the component (i.e. it is \(\pm
\boldsymbol{p_0} \).

The two signs on the first component separate the sequences into two sets. The
whole sequence is fed into the LSTM and we aim to learn to classify the
sequences. This requires remembering the first element over the length of the
sequence. Due to the large throughput of the GILR-LSTM, we would expect that it
would do better when the sequence length is large. In the original formulation
of the problem (dealing in the regime with around one hundred timesteps)
, \(p\) is set equal to \(n\). Since this would make the size of the input data
grow impracticaly large as \(\mathcal{O}(n^2)\) for long sequences, we fix \(p =
128\) and vary \(n\).

We generated three sets of data: for \(n\) equal to 1,024, 8,192, and
1,048,576. For each of these we compared a two-layer GILR-LSTM with 512 hidden
units to a canonical LSTM network with roughly the same number of parameters. We
could imagine using a two-layer LSTM with 512 hidden units in each layer, or a
one-layer LSTM with 1024 hidden units.

Initial experiments showed that the two-layer LSTM was much quicker to converge
than the one-layer LSTM, so here we compare the GILR-LSTM to the two-layer
CudnnLSTM.  We ran all experiments on a NVIDIA K80 GPU, choosing the largest
minibatch size which fit into the GPU memory, with five runs per configuration
allowing us to find the average and standard deviation of the time and number of
iterations to convergence. For all CUDA runs, a brief search over learning rate
and batch size was carried out to find the parameters which allow the network to
converge most rapidly. The criterion for convergence was five consecutive
minibatches giving 100\% accuracy. As can be seen from the learning curves
in figure \ref{fig:synthetic_training}, this was a reasonable criterion. For the
longest sequence-length, we didn't observe the CuDNN-LSTM converging.

The results illustrate that the GILR-LSTM is able to robustly outperform the
CuDNN-LSTM, which is currently the fastest LSTM implementation available. This
is somewhat surprising, as we might expect that the LSTM would do particularly
well at an example that was specifically constructed for it. We can conclude that
the nonlinearities in the LSTM's design are not vital for its performance given
that the linear GILR-LSTM is able to converge, and that if these nonlinearities are
not completely necessary, they entail a large performance cost by diallowing the
plr algorithm from being used. Furthermore, the sequence length of over one million
is the largest sequence length we have seen seriously discussed. The GILR
handles it straightforwardly.

@@

\subsection{The Medical Setting}
A common application of RNNs is in medicine. In order to illustrate the
increased performance that the plr algorithm allows, we implement an GILR-LSTM
on the problem presented in , the 2016 PhysioNet/Computing in
Cardiology Challenge 2016.  The aim of the Classification of Normal/Abnormal
Heart Sound Recordings challenge is to determine which of a set of EEGs are
produced by people with abnormal hearts.  The data is recorded from a variety of
sources and split into a training and testing set. The testing set is not
publicly available, but can be evaluated through a web portal. The top results
cover a large range of techniques, but generally rely on featurization of the
raw EEG records. We want to see if we are able to leverage our advantage at high
sequence-lenghts to train directly on the raw EEG. The EEGs are sampled at 2,000
Hz and range in length from several seconds to more than one hundred seconds,
giving a max sequence length of over 200,000. Since the testing server is not
equipped with a GPU, we cannot run PLR in the inference. We resample the
records to 500 Hz so that we are able to evaluate the test results in a
reasonable amount of time, padding all records that are shorter than the longest.

The training set of data is unbalanced, with 75\% of the data corresponding to
normal hearts. We rebalance the data by oversampling the abnormal data by a factor
of three, including each abnormal heart three times. We train a network consisting
of three stacked GIRL-LSTM layers, each with 16 hidden units. Although reasonably
small, since the input is only one-dimensional 16 hidden units allows us to
develop representations. PLR allows us to train the network very rapidly over a
long sequence length, so we don't have to downsample and lose information or
spend lots of time developing methods to form features.
We were able to fit exactly to the training set after
a few hours of training, and on the unseen test data we obtained a score of
PLACEHOLDER.

@@

\section{Conclusion}
Parallel linear recurrence is an extremely powerful algorithm and the GILR-LSTM
is just one of many possible models that can be built with it. While we were
previously forced to use large minibatches in order to obtain reasonable
training speed, with PLR in many cases we can train rapidly on small minibatch
sizes, freeing us to utilise extremely long sequence lengths and more complex
models. Our CUDA kernel is up to 40 times faster than a serial kernel on
sequence lengths of 65,536, and this benefit transfers to implemented models,
speeding up the SRU \cite{lei2017} by a factor of 9. The impressive results in
\cite{lei2017} and other recent work comprehensively show that
linearly-recurrent RNNs are, in many respects, equally as powerful as nonlinear
LSTMs. Our results on the synthetic example continue this trend, showing that
the GILR-LSTM is able to handle sequence length regimes that are longer than any
we have seen seriously suggested in the literature. We feel that the framework
of describing linear RNNs as linear surrogates for nonlinear RNNs is a fruitful
approach, and that this, paired with the PLR algorithm, will substantially
advance the ability to train RNNs, and to understand timeseries data.

>>>>

\subsection{Throughput benchmarks}

\subsubsection{Kernel performance}
We first illustrate the throughput advantage of the parallel scan algorithm for
evaluating the linear recurrence. For a minibatch comprised of \(b\) sequences
of length \(T\), we define the number of events as \(bT\) and the throughput as
the number of events per second. We implement two CUDA kernels, one which
evaluates the parallel linear recurrence described in algorithm \ref{alg:plr},
and one which evaluates the linear recurrence on GPU using the standard serial
approach (with parallelization over features but not sequence length).
This comparison is performed directly at the kernel level, avoiding
any overhead from TensorFlow. We find that the parallel kernel has
a distinct advantage at long sequence lengths with a speedup factor of up to 40x,
as shown in table \ref{table:kernel-throughput}.
At short sequence lengths, the parallel kernel does not perform well due to the
overhead of multiple passes over data and communication between processors.
\begin{table}
\begin{center}
  \begin{tabular}{@{}llll@{}}
    \label{table:kernel-throughput}
Sequence Length & \(m=4\)  & \(m=32\) & \(m=128\) \\ \midrule
16              & 0.06 & 0.06 & 0.05  \\
256             & 0.22 & 0.22 & 0.86  \\
4,096           & 1.02 & 2.94 & 3.36  \\
65,536          & 38.5 & 41.8 & 17.5  \\ \bottomrule
  \end{tabular}
\end{center}
\caption{
  Speedup of the parallel kernel compared to the serial kernel at various
  values of $(\text{minibatch size } \mathbf{x} \text{ features}) = m$.
  Performance was directly measured without any overhead from TensorFlow.
}
\end{table}

\subsubsection{Accelerating existing RNN architectures}

Many recently introduced RNNs can be accelerated with the parallel linear recurrence
algorithm.
Table \ref{table:rnn-throughput} shows the
throughput advantage from using parallel linear recurrence compared to
serial linear recurrence. Both methods compute an identical recurrence,
so using parallel rather than serial causes no numerical changes.

We controlled for GPU memory usage by choosing minibatch size such that the product
of minibatch size and sequence length is constant. We see that the simpler architectures
(for which the linear recurrence is a higher proportion of the
computational burden) are more affected by the switch to the parallel kernel.  This is
particularly clear in the case of the QRNN, where including a wider
convolutional window results in more time spent outside of the linear recurrence and
therefore less to gain from the parallelization.
\begin{table}
\begin{center}
  \begin{tabular}{@{}lrrrr@{}}
    \label{table:rnn-throughput}
Sequence Length & GILR-LSTM & SRU & QRNN(2) & QRNN(10)\\ \midrule
16 & 0.61 & 0.28 & 0.38 & 0.78\\
256 & 0.91 & 0.84 & 0.86 & 0.99\\
4,096 & 0.98 & 1.38 & 1.18 & 1.05\\
65,536 & 1.41 & 9.21 & 6.68 & 2.05\\ \bottomrule
  \end{tabular}
\end{center}
\caption{
  Speedup obtained from using parallel linear recurrence rather than
  serial linear recurrence. We analyse this for
  SRUs \cite{lei2017} and QRNNs \cite{bradbury2017quasi} with a convolutional
  filters of size 2 and 10, and GILR-LSTM which we introduce in section \ref{sec:ls-rnns}.
  We controlled for GPU memory by setting minibatch size $b$ such that
  $bT = 65536$ for sequence length $T$.
}
\end{table}

\Chris{For table \ref{table:rnn-throughput} how many layers and what was hidden size?}

\subsubsection{GILR-LSTM throughput}

Finally, we compare the throughput of the GILR-LSTM to the current fastest
LSTM implementation (which comes from the CuDNN library). Figure
\ref{fig:tp_perf} shows that the GILR-LSTM has higher throughput for most of the
small minibatch regime, with a 30x speedup at a sequence length of 65,536.
The GILR-LSTM allows us to achieve high throughput for all choices of minibatch size and
sequence length.

@@

\Chris{Put title on Figure \ref{fig:tp_perf}}

\begin{figure}[t]
\centering
\includegraphics[width=12cm]{cudnn_heatmap.png}
\caption{Throughput comparison between LSTM-256-256 with GILR-LSTM-256-256, with a
  32-dimensional input and a 2-dimensional output, for various batch sizes and
  sequence lengths.The LSTM only has a single row of data because its throughput
  is independent of sequence length. Entries are missing from the GILR-LSTM table
  because there was not enough memory on the GPU to handle such large batch
  sizes and sequences.}
\label{fig:tp_perf}
\end{figure}

@@

\subsection{Synthetic Experiment}
In order to demonstrate how PLR may be used to speed up tasks that the LSTM is
well-suited for, we tackle a pathological inference problem first introduced
in Hochreiter and Schmidhuber's initial LSTM paper \cite{hochreiter1997long}.
The input consists of sequences of length $n$ with each sequence element a randomly
chosen one-hot vector in $p$ dimensional space.
The first vector in each sequence is always the same, up to the sign of the component (i.e. it is $\pm
\boldsymbol{p_0}$). The sequential model must read in an entire sequence and then output
the sign of the first sequence element.
This sequence classification problem requires remembering the first element over the length of the
sequence and is a canonical example of the long-term dependency problem.
Due to the large throughput of the GILR-LSTM, we would expect that it
would do better when the sequence length is large. In the original formulation
of the problem (dealing in the regime with around one hundred timesteps)
, $p$ is set equal to $n$. As this would make the size of the input data
grow impracticaly large as \(\mathcal{O}(n^2)\) for long sequences, we fix \(p =
128\) and vary \(n\).

We generated sequences for $n$ equal to 1,024, 8,192, and
1,048,576. For each of these we compared a two layer GILR-LSTM with 512 hidden
units to a two layer LSTM with 512 hidden units per layer implemented by CuDNN.

We ran all experiments on a NVIDIA K80 GPU, choosing the largest
minibatch size which fit into the GPU memory with five runs per configuration
allowing us to find the average and standard deviation of the time and number of
iterations to convergence. A brief search over learning rate
and batch size was carried out to find the parameters which allow the network to
converge most rapidly for all runs. The criterion for convergence was five consecutive
minibatches giving 100\% accuracy. The learning curves
in figure \ref{fig:synthetic_training} give support to this being a reasonable
convergence criteria.
For the longest sequence length, we could not run long enough to observe the LSTM converge.

The results illustrate that the GILR-LSTM is able to converge much more quickly than
the fastest available LSTM implementation.
This is somewhat surprising given the LSTM was specifically constructed for problems
of this sort. This GILR-LSTM convergence
leads to the conclusion that the non-linearities present in
LSTM are not necessary for solving this instance of the long-term dependency problem.
The time to convergence further leads to the conclusion that inclusion of a non-linearity
at every step incurs a significant training time slowdown.
Furthermore, the GILR-LSTM is able to learn to carry dependencies over a one million
element sequence. As far as we know, this one million step sequence experiment
is the longest sequential learning problem to be handled by neural networks to date.

@@

\subsection{The Medical Setting}
Medical waveform data consists of long sequences and is a reasonable application area for RNNs.
We applied the GILR-LSTM
on the problem presented in the 2016 PhysioNet/Computing in
Cardiology Challenge 2016.  The aim of the Classification of Normal/Abnormal
Heart Sound Recordings challenge is classify electroencephalogram (EEG) recordings
as coming from a normal or abnormal heart.
The waveforms are recorded on several sets of sensors
and split into a training and testing set. The testing set is not
publicly available, but can be evaluated through a web portal. The top results
cover a large range of techniques, but generally rely on manual featurization of the
raw EEG records. The GILR-LSTM's ability to handle long sequences makes it a
candidate for training directly on the raw EEG signal.
The EEGs are sampled at 2,000 Hz
and range in length from several seconds to more than one hundred seconds,
giving a max sequence length of over 200,000. Since the testing server is not
equipped with a GPU, we cannot run PLR in the inference. We resample the
records to 500 Hz so that we are able to evaluate the test results in a
reasonable amount of time, padding all records that are shorter than the longest.

The training set of data is unbalanced, with 75\% of the data corresponding to
normal hearts. We rebalance the data by oversampling the abnormal data by a factor
of three, including each abnormal heart three times. We train a network feeding
the one dimensional input signal into three stacked GILR-LSTM layers of 16 hidden units each.
Although reasonably small, 16 hidden units allows developing
develop representations. Parallel linear recurrence allows us to rapidly train the network
over the entire long sequence length, so we don't have to work on truncated samples,
downsampled sequenes, or
spend lots of time manually developing EEG specific features.
The model obtained perfect classification on the train dataset after
a few hours \Chris{how many?} of training.
On the unseen test data we obtained a score of PLACEHOLDER.

@@

\section{Conclusion}
Parallel linear recurrence can accelerate existing RNNs and be used in the design
of new architectures such as the GILR-LSTM. Parallel linear recurrence enables
rapid training with small minibatches, which further enables training models on
extremely long sequence lengths.
Our CUDA kernel is up to 40 times faster than a serial kernel on
sequence lengths of 65,536, and this benefit transfers to implemented models,
speeding up the SRU \cite{lei2017} by a factor of 9. The impressive results in
\cite{lei2017} and other recent work comprehensively show that
linear RNNs are, in many respects, equally as powerful as nonlinear
LSTMs. Our results on the synthetic example continue this trend, showing that
the GILR-LSTM is able to handle sequence length regimes that are longer than any
we have seen seriously suggested in the literature. We feel that the framework
of describing linear RNNs as linear surrogates for nonlinear RNNs is a fruitful
approach, and that this, paired with the PLR algorithm, will substantially
advance the ability to train RNNs, and to understand timeseries data.
